---
title: "Benchmarking"
---

Benchmarking effectively refers to timing execution of our code, although aspects of memory utilisation can also be compared during benchmarking.

There are many ways to benchmark code in R and here we'll review some of the most useful ones.

## Simple benchmarking

### `system.time()`

The function takes an expression and returns the CPU time used to evaluate the expression. It is a primitive function and part of base R an a good starting point for timing R expressions.

```{r}
system.time(
    for(i in 1:100) mad(runif(1000))
    )
```

The output prints a named vector of length 3.

The first two entries are the **total user** and **system CPU** times of *the current **R** process and any child processes* on which it has waited, and the third entry is the **'real' elapsed time** since the process was started.

The resolution of the times will be system-specific.

Expressions across multiple lines can be run by enclosing them in curly braces (`{}`).

```{r}
system.time({
    for(i in 1:100) {
        v <- runif(1000)
        mad(v)
    }
})
```

### `tictoc` 📦

Package `tictoc` provides similar functionality to `system.time` with some additional useful features. To time code execution, you can wrap them between function calls `tic` and `toc`.

```{r}
library(tictoc)

tic()

for (i in 1:100) {
    v <- runif(1000)
    mad(v)
}

toc()
```

The nice features of the `tictoc` package is that;

-   Descriptive messages can be associated with each timing through `tic` argument `msg`.

-   Timings can be nested.

-   Timings can be logged through `toc` argument `log = TRUE` and accessed afterwards through `tic.log()`.

```{r}
tic(msg = "MAD 100 iterations of 1000 random values")

for (i in 1:100) {
    v <- runif(1000)
    mad(v)
}

toc(log = TRUE)

tic(msg = "MAD 1000 iterations of 1000 random values")

for (i in 1:1000) {
    v <- runif(1000)
    mad(v)
}

toc(log = TRUE)

tic(msg = "MAD 100 iterations of 10000 random values")

for (i in 1:100) {
    v <- runif(10000)
    mad(v)
}

toc(log = TRUE)

tic.log()
```

## Formal Benchmarking

`system.time()` and `tictoc` are straightforward and simple to use for timing individual of code. However, there do have a few limitations:

-   They only time execution time of the code once. As we've seen there's a lot of other things going on on your system which may affect execution time at any one time. So timings can vary if tested repeatedly.

-   Comparing different expressions has to be performed rather manually, especially if we want to also check expressions being compared give the same result.

There are a number of packages in R that make benchmarking code, especially the comparison of different expressions much easier and robust. Here we will explore two of these, `microbenchmark` and `bench`.

### `microbenchmark` 📦

The `microbenchmark()` function in package `microbenchmark` serves as a more accurate replacement of `system.time()`. To achieved this, the sub-millisecond (supposedly nanosecond) accurate timing functions most modern operating systems provide are used. This allows us to compare expressions with much shorter execution times.

Some nice package features:

-   By default evaluates each expression multiple times, by default 100 but this number can be controlled through argument `times`.

-   You can enforce checks on the results to ensure each expression tested returns the same result, with various levels of strictness through argument `check`.

-   You can supply setup code that will be run by each iteration without contributing to the timing through argument `setup`.

-   Note that the function is only meant for micro-benchmarking small pieces of source code and to compare their relative performance characteristics.

See the function documentation for more info.

Let's go ahead and look at an example.

#### Centering data in data.frame by column mean

Let's say we are given the following code and asked to speed it up.

In this example, first a data frame is created that has 151 columns. One of the columns contains a character ID, and the other 150 columns contain numeric values.

For each numeric column, the code calculates the mean and subtracts it from the values in the column, so the data in each column is centered on the mean.

```{r}
#| eval: false
rows <- 400000
cols <- 150
data <- as.data.frame(x = matrix(rnorm(rows * cols, mean = 5), ncol = cols))
data <- cbind(id = paste0("g", seq_len(rows)), data)

# Get column means
means <- apply(data[, names(data) != "id"], 2, mean)

# Subtract mean from each column
for (i in seq_along(means)) {
    data[, names(data) != "id"][, i] <- data[, names(data) != "id"][, i] - means[i]
}
```

Looking at it, we might think back to the age old R advice to "avoid loops". So to improve performance of our code we might start by working on the for loop.

Let's use `microbenchmark` to test the execution times of alternative approaches and compare them to the original approach:

-   Give we we want to iterate across more than one objects (`means` and the columns in our data) might consider the `mapply` function. We then reassign the result back to the appropriate columns in our data.frame.
-   We could also try `purrr`s `map2_dfc` which takes two inputs and column binds the results into a data.frame. We can again reassign the result back to the appropriate columns in our data.frame.

To test out our approaches we might want to make a smaller version of our data to work with. Let's create a smaller data frame with 40,000 rows and 50 columns and re-calculate our `means` vector:

```{r}
rows <- 40000
cols <- 50
data <- as.data.frame(x = matrix(rnorm(rows * cols, mean = 5), ncol = cols))
data <- cbind(id = paste0("g", seq_len(rows)), data)

means <- apply(data[, names(data) != "id"], 2, mean)
```

Now let's wrap our original for loop and our two test approaches in a benchmark.

We can wrap each expression in `{}` and pass it as a named argument for easier review of our benchmark results.

Let's only run 50 iterations instead of the default 100 through argument `times`.

Let's also include some `setup` code so that a new `data_bnch` object is created before each benchmark expression so that we don't overwrite the `data` object in the global environment.

```{r}
microbenchmark::microbenchmark(
    for_loop = {
        for (i in seq_along(means)) {
            data_bnch[, names(data_bnch) != "id"][, i] <- 
                data_bnch[, names(data_bnch) != "id"][, i] - means[i]
        }
    }, 
    mapply = {
        data_bnch[, names(data_bnch) != "id"] <- mapply(
            FUN = function(x, y) x - y,
            data_bnch[, names(data_bnch) != "id"], 
            means)
    },
      map2_dfc = {
        data_bnch[, names(data_bnch) != "id"] <- purrr::map2_dfc(
                            data_bnch[, names(data_bnch) != "id"], 
                            means, 
                            ~.x - .y)
    },
    times = 50,
    setup = {data_bnch <- data}
)
```

The results of our benchmark return one row per expression tested.

-   `expr` contains the name of the expression.

-   `min`, `lq`, `mean`, `median`, `uq` and `max` are summary statistics of the execution times across all iterations.

-   `neval` shows the numbers of iterations

So far, `purrr::map2_dfc()` is looking like the best option.

But are we sure we are getting the same results form each approach?

To ensure this we can re-run our benchmarks using the `check` argument. A value of `"equal"` will compare all values output by the benchmark using `all.equal().` For the comparison to work, we need the last expression of the computation to output the same object. As this differs in the for loop from the two others, we include a call to print the final object `data_bnch` for comparison in each expression.

```{r}
microbenchmark::microbenchmark(
    for_loop = {
        for (i in seq_along(means)) {
            data_bnch[, names(data_bnch) != "id"][, i] <- 
                data_bnch[, names(data_bnch) != "id"][, i] - means[i]
        }
        data_bnch
    }, 
    mapply = {
        data_bnch[, names(data_bnch) != "id"] <- mapply(
            function(x, y) x - y,
            data_bnch[, names(data_bnch) != "id"], 
            means)
        data_bnch
    },
      map2_dfc = {
        data_bnch[names(data_bnch) != "id"] <- purrr::map2_dfc(
                            data_bnch[, names(data_bnch) != "id"], 
                            means, 
                            ~.x - .y)
        data_bnch
    },
    times = 50,
    setup = {data_bnch <- data},
    check = "equal"
)
```

Excellent! We can now be confident that our tests are retuning the same result.

Finally, if you like to compare things visually, the output of our benchmark can be provided as an input to the `autoplot.microbenchmark` method to produce a graph of microbenchmark timings.

```{r}
#| message: false
library(ggplot2)
library(dplyr)
microbenchmark::microbenchmark(
    for_loop = {
        for (i in seq_along(means)) {
            data_bnch[, names(data_bnch) != "id"][, i] <- 
                data_bnch[, names(data_bnch) != "id"][, i] - means[i]
        }
        data_bnch
    }, 
    mapply = {
        data_bnch[, names(data_bnch) != "id"] <- mapply(
            function(x, y) x - y,
            data_bnch[, names(data_bnch) != "id"], 
            means)
        data_bnch
    },
      map2_dfc = {
          data_bnch[names(data_bnch) != "id"] <- purrr::map2_dfc(
                            data_bnch[, names(data_bnch) != "id"], 
                            means, 
                            ~.x - .y)
        data_bnch
    },
    times = 50,
    setup = {data_bnch <- data},
    check = "equal"
) %>%
autoplot()
```

### `bench` 📦

`bench` is similar to `microbenchmark`. However it offers some additional features which means I generally prefer it.

The main function equivalent to `microbenchmark()` is `mark()`.

#### `mark()`

**PROs**

-   The main pro in my view is that it also tracks memory allocations for each expression.

-   It also tracks the number and type of R garbage collections per expression iteration.

-   It verifies equality of expression results by default, to avoid accidentally benchmarking inequivalent code.

-   It allows you to execute code in separate environments (so that objects in global environment are not modified).

**Some cons to consider:**

-   it doesn't have a `setup` option.

-   the output object, while much more informative than that of `microbenchmark` can be quite bloated itself.

So let's go ahead and run our tests using `bench::mark()`.

Because there is no setup option, we need to create `data_bnch` at the start of each expression. We can also use argument `env = new.env()` to perform all our computation in a separate environment.

Because `mark()` checks for equality by default, we use the version of our expressions that print the resulting `data_bnch` at the end for comparison.

```{r}
bench::mark(
    for_loop = {
        data_bnch <- data
        for (i in seq_along(means)) {
            data_bnch[, names(data_bnch) != "id"][, i] <- 
                data_bnch[, names(data_bnch) != "id"][, i] - means[i]
        }
        data_bnch
    }, 
    mapply = {
        data_bnch <- data
        data_bnch[, names(data_bnch) != "id"] <- mapply(
            function(x, y) x - y,
            data_bnch[, names(data_bnch) != "id"], 
            means)
        data_bnch
    },
      map2_dfc = {
          data_bnch <- data
          data_bnch[names(data_bnch) != "id"] <- purrr::map2_dfc(
                            data_bnch[, names(data_bnch) != "id"], 
                            means, 
                            ~.x - .y)
        data_bnch
    },
    env = new.env()
)
```

Let's have a look at the output in detail:

-   `expression` - `bench_expr` The deparsed expression that was evaluated (or its name if one was provided).

-   `min` - The minimum execution time.

-   `median` - The sample median of execution time.

-   `itr/sec` - The estimated number of executions performed per second.

-   `mem_alloc` - Total amount of memory allocated by R while running the expression.

-   `gc/sec` - The number of garbage collections per second.

-   `n_itr` - Total number of iterations after filtering garbage collections (if `filter_gc == TRUE`).

-   `n_gc` - Total number of garbage collections performed over all iterations.

-   `total_time` - The total time to perform the benchmarks.

-   `result` - `list` A list column of the object(s) returned by the evaluated expression(s).

-   `memory` - `list` A list column with results from [`Rprofmem()`](http://127.0.0.1:30837/help/library/bench/help/Rprofmem).

-   `time` - `list` A list column of vectors for each evaluated expression.

-   `gc` - `list` A list column with tibbles containing the level of garbage collection (0-2, columns) for each iteration (rows).

I find the addition of the `mem_alloc` particularly useful. Just look at the difference between `mapply` and the other two approaches in term of memory usage!

As you can see, there's a lot more information in the `bench::mark()` output. Note as well there are a number of list columns at the which include results of the evaluated expressions, results of `Rprofmem()` and a list of garbage collection events. This can be quite useful to dig into.

However, if this object is assigned to a variable or you try to save it, it could take up A LOT of space depending on the size of the results and number of internal calls. So I recommend getting rid of such columns if you want to save benchmarks.

#### `press()`

Another cool feature of the `bench` package is bench::pressing() using the `press()` function. `press()` can be used to run [`mark()`](http://127.0.0.1:30837/help/library/bench/help/mark) across a grid of parameters and then *press* the results together.

We set the parameters we want to test across as named arguments and a grid of all possible combinations is automatically created.

The code to setup the benchmark is passed as a single unnamed expression before calling the bench::mark() code we want to run with the grid of parameters.

Let's have a look at how this works.

Say we want to test the performance of our three approaches on data.frames of different sizes varying both rows and columns.

We can specify two parameters in bench::press() as named arguments `rows` and `columns` and assigns vectors of the values we want press() to create a testing grid from.

The next curly braces `{}` contain our setup code which create data.frames of different sizes and the benchmark.

```{r}
bp <- bench::press(
    rows = c(1000, 10000, 400000),
    cols = c(10, 50, 100),
    {
        {
            # Bench press setup code:
            # create data.frames of different sizes using parameters 
            # rows & columns
            set.seed(1)
            data <- as.data.frame(x = matrix(
                rnorm(rows * cols, mean = 5), 
                ncol = cols))
            data <- cbind(id = paste0("g", seq_len(rows)), data)
            means <- apply(data[, names(data) != "id"], 2, mean)
        }
        bench::mark(
            for_loop = {
                data_bnch <- data
                for (i in seq_along(means)) {
                    data_bnch[, names(data_bnch) != "id"][, i] <- 
                        data_bnch[, names(data_bnch) != "id"][, i] - means[i]
                }
                data_bnch
            }, 
            mapply = {
                data_bnch <- data
                data_bnch[, names(data_bnch) != "id"] <- mapply(
                    function(x, y) x - y,
                    data_bnch[, names(data_bnch) != "id"], 
                    means)
                data_bnch
            },
            map2_dfc = {
                data_bnch <- data
                data_bnch[names(data_bnch) != "id"] <- purrr::map2_dfc(
                    data_bnch[, names(data_bnch) != "id"], 
                    means, 
                    ~.x - .y)
                data_bnch
            },
            env = new.env(),
            time_unit = "us"
        )
    }
)
bp
```

Now when we look at our benchmark we see we get results for each approach and also for each `row` x `column` combination.

Let's plot the results again using `autoplot` to get a better overview of our results

```{r}
autoplot(bp)
```

What thus show is that:

-    for the smallest data.frame sizes, `mapply` is actually quite performant!

-   for loops are also fastest when the number of columns is small regardless of number of rows.

-   `map2_dfc` becomes the better performer as number of columns increases.
