---
title: "Benchmarking"
---

Benchmarking effectively refers to timing execution of our code, although aspects of memory utilisation can also be compared during benchmarking.

There are many ways to benchmark code in R and here we'll review some of the most useful ones.

## Simple benchmarking

### `system.time()`

The function takes an expression and returns the CPU time used to evaluate the expression. It is a primitive function and part of base R an a good starting point for timing R expressions.

```{r}
system.time(
    for(i in 1:100) mad(runif(1000))
    )
```

The output prints a named vector of length 3.

The first two entries are the **total user** and **system CPU** times of *the current **R** process and any child processes* on which it has waited, and the third entry is the **'real' elapsed time** since the process was started.

The resolution of the times will be system-specific.

Expressions across multiple lines can be run by enclosing them in curly braces (`{}`).

```{r}
system.time({
    for(i in 1:100) {
        v <- runif(1000)
        mad(v)
    }
})
```

### `tictoc` 📦

Package `tictoc` provides similar functionality to `system.time` with some additional useful features. To time code execution, you can wrap them between function calls `tic` and `toc`.

```{r}
library(tictoc)

tic()

for (i in 1:100) {
    v <- runif(1000)
    mad(v)
}

toc()
```

The nice features of the `tictoc` package is that;

-   Descriptive messages can be associated with each timing through `tic` argument `msg`.

-   Timings can be nested.

-   Timings can be logged through `toc` argument `log = TRUE` and accessed afterwards through `tic.log()`.

```{r}
tic(msg = "MAD 100 iterations of 1000 random values")

for (i in 1:100) {
    v <- runif(1000)
    mad(v)
}

toc(log = TRUE)

tic(msg = "MAD 1000 iterations of 1000 random values")

for (i in 1:1000) {
    v <- runif(1000)
    mad(v)
}

toc(log = TRUE)

tic(msg = "MAD 100 iterations of 10000 random values")

for (i in 1:100) {
    v <- runif(10000)
    mad(v)
}

toc(log = TRUE)

tic.log()
```

## Formal Benchmarking

`system.time()` and `tictoc` are straightforward and simple to use for timing individual of code. However, there do have a few limitations:

-   They only time execution time of the code once. As we've seen there's a lot of other things going on on your system which may affect execution time at any one time. So timings can vary if tested repeatedly.

-   Comparing different expressions has to be performed rather manually, especially if we want to also check expressions being compared give the same result.

There are a number of packages in R that make benchmarking code, especially the comparison of different expressions much easier and robust. Here we will explore two of these, `microbenchmark` and `bench`.

### `microbenchmark` 📦

The `microbenchmark()` function in package `microbenchmark` serves as a more accurate replacement of `system.time()`. To achieved this, the sub-millisecond (supposedly nanosecond) accurate timing functions most modern operating systems provide are used. This allows us to compare expressions with much shorter execution times.

Some nice package features:

-   By default evaluates each expression multiple times, by default 100 but this number can be controlled through argument `times`.

-   You can enforce checks on the results to ensure each expression tested returns the same result, with various levels of strictness through argument `check`.

-   You can supply setup code that will be run by each iteration without contributing to the timing through argument `setup`.

-   Note that the function is only meant for micro-benchmarking small pieces of source code and to compare their relative performance characteristics.

See the function documentation for more info.

Let's go ahead and look at an example.

#### Centering data in data.frame by column mean

Let's say we are given the following code and asked to speed it up.

In this example, first a data frame is created that has 151 columns. One of the columns contains a character ID, and the other 150 columns contain numeric values.

For each numeric column, the code calculates the mean and subtracts it from the values in the column, so the data in each column is centered on the mean.

```{r}
#| eval: false
rows <- 400000
cols <- 150
data <- as.data.frame(x = matrix(rnorm(rows * cols, mean = 5), ncol = cols))
data <- cbind(id = paste0("g", seq_len(rows)), data)


data1 <- data   # Store in another variable for this run

# Get column means
means <- apply(data1[, names(data1) != "id"], 2, mean)

# Subtract mean from each column
for (i in seq_along(means)) {
    data1[, names(data1) != "id"][, i] <- data1[, names(data1) != "id"][, i] - means[i]
}

```

Looking at it, we might think back to the age old R advice to "avoid loops! Vectorise". So to improve performance of our code we might start by working on the for loop.

Let's use benchmark to test a few options and compare it to the original approach:

-   To vectorise, we might consider the `mapply` function, given we want to vectorise across more than one objects (`means` and the data)

```{r}

microbenchmark::microbenchmark(
    for_loop = {
        data_ctr <- data
        for (i in seq_along(means)) {
            data_ctr[, names(data_ctr) != "id"][, i] <- 
                data_ctr[, names(data_ctr) != "id"][, i] - means[i]
        }
    }, 
    mapply = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- mapply(
            function(x, y) x - y,
            data_ctr[, names(data_ctr) != "id"], 
            means)
    },
    lrg_vector = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- 
            data_ctr[, names(data_ctr) != "id"] - rep(means, each = rows)
    },
      map2_dfc = {
        data_ctr <-  cbind(data["id"],
                        purrr::map2_dfc(
                            data[, names(data) != "id"], 
                            means, 
                            ~.x - .y))
    },
    times = 5
)


```

```{r}
microbenchmark::microbenchmark(
    for_loop = {
        data_ctr <- data
        for (i in seq_along(means)) {
            data_ctr[, names(data_ctr) != "id"][, i] <- 
                data_ctr[, names(data_ctr) != "id"][, i] - means[i]
        }
    }, 
    mapply = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- mapply(
            function(x, y) x - y,
            data_ctr[, names(data_ctr) != "id"], 
            means)
    },
    lrg_vector = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- 
            data_ctr[, names(data_ctr) != "id"] - rep(means, each = times)
    },
      map2_dfc = {
        data_ctr <-  cbind(data["id"],
                        purrr::map2_dfc(
                            data[, names(data) != "id"], 
                            means, 
                            ~.x - .y))
    },
    times = 5,
    check = "equivalent"
)

```

```{r}
mb <- microbenchmark::microbenchmark(
    for_loop = {
        data_ctr <- data
        for (i in seq_along(means)) {
            data_ctr[, names(data_ctr) != "id"][, i] <- 
                data_ctr[, names(data_ctr) != "id"][, i] - means[i]
        }
        data_ctr
    }, 
    mapply = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- mapply(
            function(x, y) x - y,
            data_ctr[, names(data_ctr) != "id"], 
            means)
        data_ctr
    },
    lrg_vector = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- 
            data_ctr[, names(data_ctr) != "id"] - rep(means, each = times)
        data_ctr
    },
      map2_dfc = {
        data_ctr <-  cbind(data["id"],
                        purrr::map2_dfc(
                            data[, names(data) != "id"], 
                            means, 
                            ~.x - .y))
        data_ctr
    },
    times = 5,
    check = "equivalent"
)

mb

autoplot(mb)
```

```{r}

        data1 <- data
        for (i in seq_along(means)) {
            data1[, names(data1) != "id"][, i] <- 
                data1[, names(data1) != "id"][, i] - means[i]
        }
        data1

 
        data2 <- data
        data2[, names(data2) != "id"] <- mapply(
            function(x, y) x - y,
            data2[, names(data2) != "id"], 
            means)


        data3 <- data
        data3[, names(data3) != "id"] <- 
            data3[, names(data3) != "id"] - rep(means, each = times)

   
        data4 <-  dplyr::mutate_if(data, is.numeric, 
                                   ~scale(.x, 
                                          center = TRUE,
                                          scale = FALSE)) 
 
        data5 <-  cbind(data["id"],
                        purrr::map2_dfc(
                            data[, names(data) != "id"], 
                            means, 
                            ~.x - .y))
        
        
        
        ,
  
        
```

```{r}
waldo::compare(data1, data2)
waldo::compare(data1, data3)
waldo::compare(data3, data4)
waldo::compare(data1, data5)


all.equal(data1, data2)
all.equal(data1, data3)
all.equal(data3, data4)
all.equal(data1, data5)


```

```{r}
microbenchmark::microbenchmark(

    mapply = {
        data1 <- data
        data1[, names(data1) != "id"] <- mapply(
            function(x, y) x - y,
            data1[, names(data1) != "id"], 
            means)
        data1
    },
    lrg_vector = {
        data2 <- data
        data2[, names(data2) != "id"] <- 
            data2[, names(data2) != "id"] - rep(means, each = times)
        data2
    },
    times = 5,
    check = "equivalent"
)

```

Let's say we wanted to try and impor

### `bench` 📦

```{r}
bench::mark(
    for_loop = {
        data_ctr <- data
        for (i in seq_along(means)) {
            data_ctr[, names(data_ctr) != "id"][, i] <- 
                data_ctr[, names(data_ctr) != "id"][, i] - means[i]
        }
        data_ctr
    }, 
    mapply = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- mapply(
            function(x, y) x - y,
            data_ctr[, names(data_ctr) != "id"], 
            means)
        data_ctr
    },
    lrg_vector = {
        data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- 
            data_ctr[, names(data_ctr) != "id"] - rep(means, each = times)
        data_ctr
    },
      map2_dfc = {
        data_ctr <-  cbind(data["id"],
                        purrr::map2_dfc(
                            data[, names(data) != "id"], 
                            means, 
                            ~.x - .y))
        data_ctr
    },
    iterations = 5
)
```

```{r}
data_ctr <- data
        data_ctr[, names(data_ctr) != "id"] <- mapply(
            function(x, y) x - y,
            data_ctr[, names(data_ctr) != "id"], 
            means)
        data_ctr
```

```{r}
bp <- bench::press(
    rows = c(1000, 10000, 400000),
    cols = c(15, 75, 150),
    {
        {
            set.seed(1)
            data <- as.data.frame(x = matrix(rnorm(rows * cols, mean = 5), ncol = cols))
            data <- cbind(id = paste0("g", seq_len(rows)), data)
            means <- apply(data[, names(data) != "id"], 2, mean)
        }
        bench::mark(
            for_loop = {
                data_ctr <- data
                for (i in seq_along(means)) {
                    data_ctr[, names(data_ctr) != "id"][, i] <- 
                        data_ctr[, names(data_ctr) != "id"][, i] - means[i]
                }
                data_ctr
            }, 
            mapply = {
                data_ctr <- data
                data_ctr[, names(data_ctr) != "id"] <- mapply(
                    function(x, y) x - y,
                    data_ctr[, names(data_ctr) != "id"], 
                    means)
                data_ctr
            },
            lrg_vector = {
                data_ctr <- data
                data_ctr[, names(data_ctr) != "id"] <- 
                    data_ctr[, names(data_ctr) != "id"] - rep(means, each = rows)
                data_ctr
            },
            map2_dfc = {
                data_ctr <-  cbind(data["id"],
                                   purrr::map2_dfc(
                                       data[, names(data) != "id"], 
                                       means, 
                                       ~.x - .y))
                data_ctr
            },
            iterations = 5,
            time_unit = "us"
        )
    }
)
```

```{r}
ggplot2::autoplot(bp)
```
