---
title: "setup"
---

## Install a recent R version

You can download the latest version of R from <https://www.r-project.org/>. Ideally you should have an R version \> 4.0.0. Otherwise you might experience issues with using/installing some of the packages needed.

## Install RStudio

Install Rstudio desktop from <https://posit.co/download/rstudio-desktop/>

## Install R dependencies

Use the following code to install the packages we will be working with in one go.

```{r}
#| eval: false
pkgs <- c(
    "arrow", 
    "beeswarm",
    "bench", 
    "benchmarkme", 
    "biglm", 
    "data.table", 
    "DBI", 
    "dbplyr", 
    "doFuture",
    "dplyr", 
    "dtplyr", 
    "duckdb", 
    "float", 
    "fs", 
    "fst", 
    "furrr",
    "future",
    "future.apply",
    "gt", 
    "ggplot2",
    "lobstr", 
    "microbenchmark", 
    "pryr", 
    "purrr",
    "qs", 
    "readr", 
    "RSQLite",
    "skimr",
    "tibble",
    "tictoc", 
    "usethis", 
    "vroom"
)

install.packages(pkgs)
```

If you are having problems installing the `arrow` package, have a look at the [package installation instructions](https://arrow.apache.org/docs/r/#installation).

### Download course materials

The course materials can be found at <https://github.com/r-rse/optimise-r-materials>.

To download the materials you can use `usethis` function `use_course()`. This will download a `zip` file of the repository and unzip it for you. If you do not supply a `destdir` path, as below, it will ask you whether you want it to download it to your `Desktop` by default. If you prefer it to be downloaded to another destination, just supply a path to `destdir`.

```{r}
#| eval: false
usethis::use_course("r-rse/optimise-r-materials")
```
